---
title: 'Chapter 6: Linear Model Selection and Regularisation'
subtitle: 'An Introduction to Statistical Learning with Applications in R'
author: "Alfredo Hernández"
date: "10 April 2018"
output:
  html_notebook:
    toc: yes
    toc_depth: 3
---

```{r message=FALSE, include=FALSE}
library(tidyverse)
source("../libs/multiplot.R")
```

# Exercise 8
	
In this exercise, we will generate simulated data, and will then use this data to perform best subset selection.

(a) Use the `rnorm()` function to generate a predictor $X$ of length $n = 100$, as well as a noise vector $\epsilon$ of length $n = 100$.
 
(b) Generate a response vector $Y$ of length $n = 100$ according to the model
$$
Y = \beta_{0} + \beta_{1}X + \beta_{2}{X}^{2} + \beta_{3}{X}^{3} + \epsilon,
$$
 
where $\beta_{0}$, $\beta_{1}$, $\beta_{2}$, and $\beta_{3}$ are constants of your choice.
 
(c) Use the `regsubsets()` function to perform best subset selection in order to choose the best model containing the predictors $X, {X}^{2}, \dots, {X}^{10}$. What is the best model obtained according to $C_{p}$, $BIC$, and adjusted $R^{2}$? Show some plots to provide evidence for your answer, and report the coefficients of the best model obtained.
 
(d) Repeat (c), using forward stepwise selection and also using backwards stepwise selection. How does your answer compare to the results in (c)?
 
(e) Now fit a Lasso model to the simulated data, again using $X, {X}^{2}, \dots, {X}^{10}$ as predictors. Use cross-validation to select the optimal value of $\lambda$. Create plots of the cross-validation error as a function of $\lambda$. Report the resulting coefficient estimates, and discuss the results obtained.
 
(f) Now generate a response vector $Y$ according to the model

$$
Y = \beta_{0} + \beta_{7} {X}^{7} + \epsilon,
$$  
 
and perform best subset selection and the Lasso. Discuss the results obtained.

<!-- ## Answers -->

## Exercise 8 (a)
To be consistent with our results, we need to set a seed first:
```{r}
set.seed(1)
```

Now we define the predictor $X$ and the noise $\epsilon$:
```{r}
X <- rnorm(100)
epsilon <- rnorm(100)
```

## Exercise 8 (b)

First of all, we define our $\beta$ coefficients:
```{r}
beta0 <- 2
beta1 <- 3
beta2 <- -2
beta3 <- -0.3
```

Now we define our $Y$ response:
```{r}
Y <- beta0 + beta1 * X + beta2 * X^2 + beta3 * X^3 + epsilon
```

We can see our function below:
```{r}
ggplot(data = tibble(X, Y)) +
	geom_point(aes(x = X, y = Y))
```

## Exercise 8 (c)

We need the `leaps` package to use the `regsubsets()` function:
```{r message=FALSE}
library(leaps)
```

Now we use `regsubsets()` to select best model having a polynomial of $X$ of degree 10:
```{r}
mod.full <- regsubsets(Y ~ poly(X, 10, raw = T), 
											 data = tibble(Y, X), 
											 nvmax = 10)
```
Note that we use the `poly()` function with the `raw = TRUE` option to use raw polynomials instead of orthogonal ones. 

Now we can get the summary of the best selection using an exhaustive algorithm (the default method of `regsubsets()`):
```{r}
(mod.summary <- summary(mod.full))[["outmat"]]
```

Now we find the model or subset size for the best $C_{p}$, $BIC$, and adjusted $R^{2}$ coefficients:
```{r}
min.cp <- which.min(mod.summary$cp)
min.bic <- which.min(mod.summary$bic)
max.adjr2 <- which.max(mod.summary$adjr2)
```

Let us remember that the previous coefficients are defined as follows:
$$
C_{p} = \frac{1}{n} (\text{RSS} + 2d \hat{\sigma}^{2}), \quad
BIC = \frac{1}{n} (\text{RSS} + \log(n) d \hat{\sigma}^{2}), \quad
\text{Adj} R^{2} = 1 - \frac{\text{RSS}/(n-d-1)}{\text{TSS}/(n-1)},
$$
where $\text{TSS} = \sum (y_{i} - \bar{y})^{2}$ is the total sum of squares.

Now we define a `ggplot2` function to simplify the process of exploring the results:
```{r}
plot_statistics <- function(summary, stat, opt.stat, title.str = NULL) {
	variable <- summary[[tolower(stat)]]
	
	gg <- ggplot(tibble(x = seq_along(variable), y = variable)) +
		geom_line(aes(x = x, y = y)) +
		geom_point(aes(x = opt.stat, y = variable[opt.stat]), colour = "red") +
		scale_x_continuous(breaks = seq(1:10)) +
		labs(x = "Subset Size", y = paste("Best Subset of", stat))
	
	if (!is.null(title.str)) {
		gg <- gg + labs(title = title.str)
	}
	
	return(gg)
}
```

Now we compare the three statistics as a function of the subset size:
```{r}
ggmatrix(list(plot_statistics(mod.summary, "Cp", min.cp),
							plot_statistics(mod.summary, "BIC", min.bic),
							plot_statistics(mod.summary, "AdjR2", max.adjr2)),
				 nrow = 3, ncol = 1,
				 yAxisLabels = c("Cp", "BIC", "AdjR2"),
				 xlab = "Subset Size",
				 ylab = "Best Subset of",
				 switch = "both") +
	theme(strip.background = element_rect(fill = "white"),
				strip.placement = "outside")
```


As we can see from the results, with $C_{p}$, $BIC$, and adjusted $R^{2}$ criteria, `r min.cp`, `r min.bic`, `r max.adjr2 `-variable models are picked, respectively. The expression for this `r min.cp`-variable and `r min.bic`-variable model are the following, respectively:
```{r}
# 4-variable model
coefficients(mod.full, id = min.cp)
# 3-variable model
coefficients(mod.full, id = min.bic)
```

As we can see, the $BIC$ statistic selects the correct $\beta_{0} + \beta_{1}X + \beta_{2}{X}^{2} + \beta_{3}{X}^{3}$ model, whilst the $C_{p}$ and adjusted $R^{2}$ statistics additionally pick $X^{5}$.

## Exercise 8 (d)
First of all, we need to fit forward and backward stepwise models to the data using the `regsubsets()` specifying the desired method:
```{r}
mod.fwd <- regsubsets(Y ~ poly(X, 10, raw = T), 
											data = tibble(Y, X), 
											nvmax = 10,
											method = "forward")
mod.bwd <- regsubsets(Y ~ poly(X, 10, raw = T), 
											data = tibble(Y, X), 
											nvmax = 10,
											method = "backward")

```

Now we can get the summary of the best selection using both models:
```{r}
(fwd.summary <- summary(mod.fwd))[["outmat"]]
(bwd.summary <- summary(mod.bwd))[["outmat"]]
```

As we did before, we find the model or subset size for the best $C_{p}$, $BIC$, and adjusted $R^{2}$ coefficients:
```{r}
min.cp.fwd <- which.min(fwd.summary$cp)
min.bic.fwd <- which.min(fwd.summary$bic)
max.adjr2.fwd <- which.max(fwd.summary$adjr2)
```
```{r}
min.cp.bwd <- which.min(bwd.summary$cp)
min.bic.bwd <- which.min(bwd.summary$bic)
max.adjr2.bwd <- which.max(bwd.summary$adjr2)
```

```{r}
ggmatrix(list(# Forward Stepwise
							plot_statistics(fwd.summary, "Cp", min.cp.fwd),
							plot_statistics(fwd.summary, "BIC", min.bic.fwd),
							plot_statistics(fwd.summary, "AdjR2", max.adjr2.fwd),
							# Backward Stepwise
							plot_statistics(bwd.summary, "Cp", min.cp.bwd),
							plot_statistics(bwd.summary, "BIC", min.bic.bwd),
							plot_statistics(bwd.summary, "AdjR2", max.adjr2.bwd)),
				 byrow = F,
				 nrow = 3, ncol = 2,
				 yAxisLabels = c("Cp", "BIC", "AdjR2"),
				 xAxisLabels = c("Forward Stepwise", "Backward Stepwise"),
				 xlab = "Subset Size",
				 ylab = "Best Subset of",
				 switch = "both") +
	theme(strip.background = element_rect(fill = "white"),
				strip.placement = "outside")
```

As we can see from the results, using the forward and backward stepwise methods we get the same exact results: the $C_{p}$ and adjusted $R^{2}$ statistics pick $X^{5}$ additionally to $X^{0}$, $X^{1}$, $X^{2}$, and $X^{3}$ whilst for the $BIC$ selects the correct model.

## Exercise 8 (e)

We need the `glmnet` package to use Lasso on the data:
```{r message=FALSE}
library(glmnet)
```

```{r}
xmat <- model.matrix(Y ~ poly(X, 10, raw = T), 
										 data = tibble(Y,X))[, -1]
mod.lasso <- cv.glmnet(xmat, Y, alpha = 1)
```
Notice that we use `[, -1]` to get rid of the intercept column.

We can easily find out the optimal $\lambda$:
```{r}
best.lambda <- mod.lasso$lambda.min
```

We can see the Lasso model in the following plot:
```{r}
plot(mod.lasso)
```


Now, the final step is to use the `predict()` function with the Lasso model to find the coefficients:
```{r}
predict(mod.lasso, s = best.lambda, type="coefficients")
```

Let us notice that Lasso predicts $\beta_{0} + \beta_{1}X + \beta_{2}{X}^{2} + \beta_{3}{X}^{3} + \beta_{10}{X}^{10}$, but $\beta_{10}$ is almost negligible.

## Exercise 8 (f)

First of all, we create a new $Y$ response with different $\beta_{7} = 7$:
```{r}
beta7 <- 7
Y2 <- beta0 + beta7 * X^7 + epsilon
```

We can see our function below:
```{r}
ggplot(data = tibble(X, Y2)) +
	geom_point(aes(x = X, y = Y2))
```

Now we use `regsubsets()` to select best model and perform a summary of the selected models:
```{r}
mod2.full <- regsubsets(Y2 ~ poly(X, 10, raw = T), 
												data = tibble(Y2, X), 
												nvmax = 10)
(mod2.summary <- summary(mod2.full))[["outmat"]]
```

Let us find the model for the best $C_{p}$, $BIC$, and adjusted $R^{2}$ coefficients:
```{r}
min2.cp <- which.min(mod2.summary$cp)
min2.bic <- which.min(mod2.summary$bic)
max2.adjr2 <- which.max(mod2.summary$adjr2)
```

```{r}
ggmatrix(list(plot_statistics(mod2.summary, "Cp", min2.cp),
							plot_statistics(mod2.summary, "BIC", min2.bic),
							plot_statistics(mod2.summary, "AdjR2", max2.adjr2)),
				 nrow = 3, ncol = 1,
				 yAxisLabels = c("Cp", "BIC", "AdjR2"),
				 xlab = "Subset Size",
				 ylab = "Best Subset of",
				 switch = "both") +
	theme(strip.background = element_rect(fill = "white"),
				strip.placement = "outside")
```

```{r}
# 2-variable model
coefficients(mod2.full, id = min2.cp)
# 1-variable model
coefficients(mod2.full, id = min2.bic)
# 4-variable model
coefficients(mod2.full, id = max2.adjr2)
```
What we see from the previous results is that $BIC$ statistic picks the most accurate `r min2.bic`-variable model with matching coefficients, whilst other criteria pick additional variables.

Let us use now use Lasso to find the best model:
```{r}
xmat2 <- model.matrix(Y2 ~ poly(X, 10, raw = T), 
										 data = tibble(Y2, X))[, -1]
mod2.lasso <-  cv.glmnet(xmat2, Y2, alpha = 1)
```

Now we find the optimal lambda:
```{r}
best2.lambda <- mod2.lasso$lambda.min
```

```{r}
best.model2 <- glmnet(xmat2, Y2, alpha = 1)
predict(best.model2, s = best.lambda, type = "coefficients")
```
As we can see, Lasso also picks the best 1-variable model, but the intercept coefficient does not give a good estimation of the real value of $\beta_{0}$. 

As a conclusion, it seems that $BIC$ and Lasso effectively select the 1-variable model, but $BIC$ gives a remarkably better estimation of $\beta_{0}$ and $\beta_{7}$.
